---
title: "Cousera Parctical Machine Learning Project"
author: "Rajesh Gore"
date: "Sunday, January 31, 2016"
output: html_document
---
## Prediction Assignment

### Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

#### Required packages
```{r, message=FALSE}
require(caret)
require(corrplot)
require(stats)
require(knitr)
require(ggplot2)
knitr::opts_chunk$set(cache=TRUE)
```

####Data
The training data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

###Prepare the datasets
```{r}
library(data.table)
setInternet2(TRUE)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
Data <- fread(url)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
DataTest <- fread(url)
```

Which variables in the test dataset have zero NAs? Use this tip: finding columns with all missing values in r.

Belt, arm, dumbbell, and forearm variables that do not have any missing values in the test dataset will be predictor candidates.
Subset the primary dataset to include only the predictor candidates and the outcome variable, classe.
Make classe into a factor.

```{r}
isAnyMissing <- sapply(DataTest, function (x) any(is.na(x) | x == ""))
isPredictor <- !isAnyMissing & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(isAnyMissing))
predCandidates <- names(isAnyMissing)[isPredictor]
predCandidates
varToInclude <- c("classe", predCandidates)
Data <- Data[, varToInclude, with=FALSE]
dim(Data)
names(Data)
Data <- Data[, classe := factor(Data[, classe])]
Data[, .N, classe]
```
Split the dataset into a 60% training and 40% probing dataset.


```{r}
seed <- as.numeric(as.Date("2016-01-31"))
set.seed(seed)
inTrain <- createDataPartition(Data$classe, p=0.6)
DTrain <- Data[inTrain[[1]]]
DProbe <- Data[-inTrain[[1]]]
```
Preprocess the prediction variables by centering and scaling.
```{r}
X <- DTrain[, predCandidates, with=FALSE]
preProc <- preProcess(X)
preProc
XCS <- predict(preProc, X)
DTrainCS <- data.table(data.frame(classe = DTrain[, classe], XCS))
##Apply the centering and scaling to the probing dataset.
X <- DProbe[, predCandidates, with=FALSE]
XCS <- predict(preProc, X)
DProbeCS <- data.table(data.frame(classe = DProbe[, classe], XCS))
#Check for near zero variance.
nzv <- nearZeroVar(DTrainCS, saveMetrics=TRUE)
if (any(nzv$nzv)) nzv else message("No variables with near zero variance")
```

Examine groups of prediction variables.

```{r}
histGroup <- function (data, regex) {
  col <- grep(regex, names(data))
  col <- c(col, which(names(data) == "classe"))
  library(reshape2)
  n <- nrow(data)
  DMelted <- melt(data[, col, with=FALSE][, rownum := seq(1, n)], id.vars=c("rownum", "classe"))
  library(ggplot2)
  ggplot(DMelted, aes(x=classe, y=value)) +
    geom_violin(aes(color=classe, fill=classe), alpha=1/2) +
#     geom_jitter(aes(color=classe, fill=classe), alpha=1/10) +
#     geom_smooth(aes(group=1), method="gam", color="black", alpha=1/2, size=2) +
    facet_wrap(~ variable, scale="free_y") +
    scale_color_brewer(palette="Spectral") +
    scale_fill_brewer(palette="Spectral") +
    labs(x="", y="") +
    theme(legend.position="none")
}
histGroup(DTrainCS, "belt")
```

```{r}
histGroup(DTrainCS, "[^(fore)]arm")
```

```{r}
histGroup(DTrainCS, "dumbbell")
```

```{r}
histGroup(DTrainCS, "forearm")
```

###Train a prediction model
Using random forest, the out of sample error should be small. The error will be estimated using the 40% probing sample. I would be quite happy with an error estimate of 3% or less.

Set up the parallel clusters.
```{r}
library(parallel)
library(doParallel)

cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
#Set the control parameters.
ctrl <- trainControl(classProbs=TRUE,
                     savePredictions=TRUE,
                     allowParallel=TRUE)
#Fit model over the tuning parameters.
method <- "rf"
system.time(trainingModel <- train(classe ~ ., data=DTrainCS, method=method))
#Stop the clusters.
stopCluster(cl)
```
###Evaluate the model on the training dataset


```{r}
trainingModel
hat <- predict(trainingModel, DTrainCS)
confusionMatrix(hat, DTrain[, classe])
```
###Evaluate the model on the probing dataset
```{r}
hat <- predict(trainingModel, DProbeCS)
confusionMatrix(hat, DProbeCS[, classe])
```
###Display the final model

```{r}
varImp(trainingModel)
trainingModel$finalModel
```
The estimated error rate is less than 1%.

Save training model object for later.

```{r}
save(trainingModel, file="trainingModel.RData")
```
###Predict on the test data
Load the training model.
```{r}
load(file="trainingModel.RData", verbose=TRUE)
#Get predictions and evaluate.
DTestCS <- predict(preProc, DataTest[, predCandidates, with=FALSE])
hat <- predict(trainingModel, DTestCS)
DataTest <- cbind(hat , DataTest)
subset(DataTest, select=names(DataTest)[grep("belt|[^(fore)]arm|dumbbell|forearm", names(DataTest), invert=TRUE)])
```

###Submission to Coursera
Write submission files to predictionAssignment_files/answers.

```{r}
pml_write_files = function(x){
  n = length(x)
  path <- "predictionAssignment_files/answers"
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(path, filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(hat)
```

